### 2019-08-03 paper/intro.tex
some more edits

\subsection{Gyrochronology}
It has long been known that the rotation periods of FGKM dwarfs increase over
time \citep{skumanich1972}.
This characteristic of main-sequence stars allows them to be dated via their
rotation periods in a practice known as gyrochronology, which is useful
since the ages of main-sequence stars are extremely difficult to measure via
the traditional age-dating method of isochrone placement.
It is also well established that stars of the same age but different {\it
masses} have different rotation periods \racomment{(citation)}, thought to be
caused by the deeper convective zones, and therefore stronger magnetic dynamos
(and more efficient magnetic braking) in lower-mass stars
\footnote{Color is the directly observable quantity to which empirical
gyrochronology relation are usually calibrated to, rather than mass or
effective temperature.}.
However, an underlying assumption behind many of the empirical gyrochronology
relations \citep[\eg][]{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015, angus2019} is that the relationships between rotation period and
color, and rotation period and age are {\it separable}.
In other words, the period-color relation is the same at all ages and the
period-age relation is the same at all colors.
It was recently shown that old stars rotate more rapidly than a simple,
separable gyrochronology relation would predict \citep{angus2015,
vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
modification to the classical \citet{skumanich1972} spin-down law is required
to reproduce the data \citep{vansaders2016, vansaders2018}.
An even more recent analysis of middle-aged open clusters provides further
indication that the exponent of the period-age relation is mass-dependent.
New rotation period measurements of the 1.1 Gyr open cluster, NGC 6811,
reveal a flattened relationship between rotation period and color
\citep{curtis2019}.
In this cluster, the G dwarfs rotate at the same rate as the K dwarfs.

Unfortunately, the open clusters with rotation period measurements are mostly
young -- currently the oldest is 2.5 Gyr (NGC 6819).
Without rotation periods for precisely dated old stars, it is extremely
difficult to calibrate the relationship between rotation period and color at
old ages.
For this reason, we turned to a population-based stellar age indicator,
velocity dispersion, to investigate the period-temperature relations at old
ages.
% Since a separable period-age and period-color relation provided an adequate
% fit for the young stars, it was naturally assumed to apply to old stars too.

@@

In this paper, we tested the \citet{angus2019} gyrochronology relation, which
was calibrated using the period-color relation of Praesepe (in \gaia\ \gcolor\
color) and the period-age relation of Praesepe and the Sun.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
Although only calibrated using Praesepe and the Sun, this gyrochronology
relation was tested on NGC 6819, a 2.5 Gyr cluster, for which it predicted
accurate ages.
In addition, the period-color relation of Praesepe is nearly identical to the
period-color relation of the Hyades, a cluster of around the same age: $\sim$
650 years \citep{douglas2016, douglas2017, rebull2017}.
However, this relation does {\it not} provide a good fit to NGC 6811, a 1.1
Gyr cluster \citep{curtis2019}.
The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
appears as though the K stars have `stalled' -- their spin-down has been
halted.
The \citet{angus2019} gyrochronology relation assumes that the rotation
period-color relation of Praesepe is applicable to stars of all ages: the same
polynomial relation fit to Praesepe is used to describe the period-color
relation for all stars.
However, the NGC 6811 cluster suggests that this assumption does not hold past
around 1 Gyr.
If the period-color relation of the \citet{angus2019} gyrochronology model
{\it were} a perfect model for the rotational evolution of stars, then groups
of stars selected to be similar ages using this relation should fall on an
isochrone on a CMD, an isochrone in period-color space, and an
isochrone in velocity dispersion-color space.
Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
variations in metallicity and extinction, blur out the main sequence enough
that differences between stars in different age and rotation bins are not
easily discernable on the CMD, although there is still a general age and
rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
\ref{fig:age_gradient}.
For this reason, we chose to use {\it kinematic} `isochrones', instead of
magnitude and color-based isochrones: although kinematics as an age indicator
is not necessarily as well calibrated to an absolute age scale as CMD
position, it can be very sensitive to {\it differences} in the ages of stellar
populations.

Isochrones and stellar evolution tracks are highly dependent on choices made
about input physics and assumptions and have often been calibrated using
different types of stars.
As a result, different sets of models can have very different shapes on the
CMD, particularly at low masses.
Instead of relying on CMD position to age-date groups of stars, we opted to
explore age trends via kinematics.
Kinematic age-dating has the advantage of being relatively model independent,
or at least, having a very simple model: that velocity dispersion increases
over time.
This means that it is relatively easy to rank groups of stars by age: older
groups have a larger velocity dispersion.
It is less easy to rank-order groups of stars by age on the CMD because
stellar evolution is highly mass and metallicity dependent, and because
different stellar evolution models predict different ages for the same CMD
position.
This does not mean that it is not possible to calibrate gyrochronology using
isochrones, however it is not the focus of this study.

>>>>

\subsection{Gyrochronology}
It has long been known that the rotation periods of FGKM dwarfs increase over
time \citep{skumanich1972}.
This characteristic of main-sequence stars allows them to be dated via their
rotation periods in a practice known as gyrochronology, which is useful
since the ages of main-sequence stars are extremely difficult to measure via
the traditional age-dating method of isochrone placement.
It is also well established that stars of the same age but different {\it
masses} have different rotation periods \racomment{(citation)}, thought to be
caused by the deeper convective zones, and therefore stronger magnetic dynamos
(and more efficient magnetic braking) in lower-mass stars.
However, an underlying assumption behind many empirical gyrochronology
relations \citep[\eg][]{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015, angus2019} is that the relationships between rotation period and
photometric color\footnote{As the directly observable quantity, color is
usually used as a mass-proxy, and empirical gyrochronology relations are
usually calibrated in color, rather than mass or effective temperature.
}, and rotation period and age are {\it separable}.
In other words, the period-color relation is the same at all ages and the
period-age relation is the same at all colors.
It was recently shown that old field stars rotate more rapidly than a simple,
separable gyrochronology relation would predict \citep{angus2015,
vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
modification to the classical \citet{skumanich1972} spin-down law is required
to reproduce the data \citep{vansaders2016, vansaders2018}.
An even more recent analysis of middle-aged open clusters provides further
indication that the exponent of the period-age relation is mass-dependent.
New rotation period measurements of the 1.1 Gyr open cluster, NGC 6811,
reveal a flattened relationship between rotation period and color
\citep{curtis2019}.
In this cluster, the G dwarfs rotate at the same rate as the K dwarfs.

Unfortunately, the open clusters with rotation period measurements are mostly
young -- currently the oldest is 2.5 Gyr (NGC 6819).
Without rotation periods for precisely dated old stars, it is extremely
difficult to calibrate the relationship between rotation period and color at
old ages.
For this reason, we used a population-based stellar age indicator, velocity
dispersion, to investigate the period-color relations at old ages.
% Since a separable period-age and period-color relation provided an adequate
% fit for the young stars, it was naturally assumed to apply to old stars too.

@@

In this paper, we tested the \citet{angus2019} gyrochronology relation, which
was calibrated using the period-color relation of Praesepe (in \gaia\ \gcolor\
color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it the ideal cluster for calibrating the period-color
relation of stars at 650 Myrs.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
Although only calibrated using Praesepe ($\sim$ 650 Myr) and the Sun (4.56
Gyr), the \citet{angus2019} gyrochronology relation predicts accurate ages for
members of NGC 6819, a 2.5 Gyr open cluster.
However there are no rotation periods for K or M dwarfs in this cluster -- its
coolest members with rotation periods are G dwarfs.
The period-color relation of Praesepe is nearly identical to the period-color
relation of the Hyades, a cluster of around the same age: $\sim$ 650 years
\citep{douglas2016, douglas2017, rebull2017}.
However, Praesepe and the Hyades do {\it not} have the same period-color
relation as NGC 6811, a 1.1 Gyr cluster \citep{curtis2019}.
The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
appears as though the K stars have `stalled' -- their spin-down has been
halted.
The \citet{angus2019} gyrochronology relation assumes that the rotation
period-color relation of Praesepe is applicable to stars of all ages: the same
polynomial relation fit to Praesepe is used to describe the period-color
relation for all stars.
However, the NGC 6811 cluster suggests that this assumption does not hold for
K dwarfs past around 1 Gyr.
If the period-color relation of the \citet{angus2019} gyrochronology model
{\it were} a perfect model for the rotational evolution of stars, then groups
of stars selected to be similar ages using this relation should fall on an
isochrone on a CMD.
% and have the same velocity dispersion across all colors.
Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
variations in metallicity and extinction, blur out the main sequence enough
that differences between stars in different age and rotation bins are not
easily discernable on the CMD, although there is still a general age and
rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
\ref{fig:age_gradient}.
For this reason, we chose to use {\it kinematic} `isochrones', instead of
magnitude and color-based isochrones: although kinematics as an age indicator
is not necessarily as well calibrated to an absolute age scale as CMD
position, it can be very sensitive to {\it differences} in the ages of stellar
populations.

Isochrones and stellar evolution tracks are highly dependent on choices made
about input physics and assumptions.
% and have often been calibrated using
% different types of stars.
As a result, different sets of models can have very different shapes on the
CMD, particularly at low masses.
In fact, only empirical models, not physical ones, are currently able to
reproduce the CMD positions of M dwarfs.
Instead of relying on CMD position to age-date groups of stars, we opted to
explore age trends via kinematics.
Kinematic age-dating has the advantage of being relatively model independent,
or at least, having a very simple model: that velocity dispersion increases
over time.
This means that it is relatively easy to rank groups of stars by age: older
groups have a larger velocity dispersion.
% It is less easy to rank-order groups of stars by age on the CMD because
% stellar evolution is highly mass and metallicity dependent, and because
% different stellar evolution models predict different ages for the same CMD
% position.
% This does not mean that it is not possible to calibrate gyrochronology using
% isochrones, however it is not the focus of this study.

### 2019-08-03 paper/method.tex
tidying up and refining text
some more edits

\begin{figure}
  \caption{
A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
    extinction-corrected magnitudes, colored by their rotation periods.
We excluded photometric binaries and subgiants from our analysis by removing
stars above the two dashed lines.
% The rotation periods of binaries and subgiants do not follow a Skumanich-like
% braking law.
% The rotation period gradient across the main sequence is visible by eye in
%     this figure: young, rapidly rotating stars are located below the old,
%     slowly rotating stars.
% since we rely on a simple scaling
% between rotation period and age to make the argument that the rotation period
% is not caused by incorrect period measurements.
}
  \centering
    \includegraphics[width=1\textwidth]{CMD_cuts}
\label{fig:CMD_cuts}
\end{figure}
To explore the age of this stellar population from a rotation-period
standpoint, it was first necessary to remove visual binaries and subgiants
from the sample.
The rotation-period evolution of these two types of stars is generally
different to that of single stars which more usually follow a Skumanich-like
spin-down law.
Tidal and magnetic interactions between the two components of a binary system
can influence the rotation periods of both stars, and the expanding envelopes
of subgiants drive rapid spin-down through conservation of angular momentum.
We removed visual binaries and subgiants from the sample by applying cuts to
the color-magnitude diagram (CMD), shown in figure \ref{fig:CMD_cuts}.
We fit a 6th-order polynomial to the main sequence and raised it by 0.22 dex,
to approximate the division between single stars and visual binaries.
We removed all stars above this line from the sample.
We removed subgiants by eliminating stars brighter than 6th magnitude in
\gaia\ G-band from the sample.

We removed stars with negative parallaxes and parallax signal-to-noise ratios
below 10 from the sample.
We also removed a small number of stars fainter than 16th magnitude from the
sample.
We used the {\tt Pyia} {\it Python} package to calculate stellar velocities.
{\tt Pyia} has built-in functionality for calculating velocity samples from
the full \gaia\ uncertainty covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the {\it covariance}
between these properties.
We removed stars with absolute \vb\ uncertainties greater than 1 \kms\ from
the sample.

>>>>

\begin{figure}
  \caption{
A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
    extinction-corrected magnitudes, colored by rotation period.
We excluded photometric binaries and subgiants from our analysis by removing
stars above the two dashed lines.
% The rotation periods of binaries and subgiants do not follow a Skumanich-like
% braking law.
% The rotation period gradient across the main sequence is visible by eye in
%     this figure: young, rapidly rotating stars are located below the old,
%     slowly rotating stars.
% since we rely on a simple scaling
% between rotation period and age to make the argument that the rotation period
% is not caused by incorrect period measurements.
}
  \centering
    \includegraphics[width=1\textwidth]{CMD_cuts}
\label{fig:CMD_cuts}
\end{figure}
To explore the age of this stellar population from a rotation-period
standpoint, it was first necessary to remove visual binaries and subgiants
from the sample.
The rotation-period evolution of these two types of stars is generally
different to that of single stars which more usually follow a Skumanich-like
spin-down law.
% Tidal and magnetic interactions between the two components of a binary system
% can influence the rotation periods of both stars, and the expanding envelopes
% of subgiants drive rapid spin-down through conservation of angular momentum.
We removed visual binaries and subgiants from the sample by applying cuts to
the color-magnitude diagram (CMD), shown in figure \ref{fig:CMD_cuts}.
We fit a 6th-order polynomial to the main sequence and raised it by 0.22 dex,
to approximate the division between single stars and visual binaries.
We removed all stars above this line from the sample.
We also removed subgiants by eliminating stars brighter than 6th magnitude in
\gaia\ G-band.
\begin{figure}
  \caption{
Dereddened main sequence stars with \mct\ rotation periods on the \gaia\ CMD
    with visual binaries removed.  Points are colored by their gyrochronal
    age, according to the
    \citet{angus2019} gyrochronology relation.
    A general age gradient is visible across the main sequence.
}
  \centering
    \includegraphics[width=1\textwidth]{age_gradient}
\label{fig:age_gradient}
\end{figure}

We removed stars with negative parallaxes and parallax signal-to-noise ratios
below 10 from the sample.
We also removed a small number of stars fainter than 16th magnitude.
We used the {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy}
\citep{astropy2013, astropy2018} {\it Python} packages to calculate stellar
velocities.
{\tt Pyia} has built-in functionality for calculating velocity samples from
the full \gaia\ uncertainty covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the {\it covariance}
between these properties.
We removed stars with absolute \vb\ uncertainties greater than 1 \kms\ from
the sample.

### 2019-08-03 paper/results.tex
some more edits

\begin{figure}
  \caption{
Top: rotation period vs effective temperature for stars in the \mct\
    catalog.
    The full catalog, with subgiants and visual binaries removed is shown in
    grey, and stars selected to be in different age groups are overlayed in
    color.
    These age groups were selected using the \citet{angus2019} gyrochronology
    relation.
The legend in the center of the figure lists the age range of each group.
    Bottom: velocity dispersion vs effective temperature for each age group.
The color of the line corresponds to the color of the group shown in the top
    panel.
If the gyrochronal model were correct at all ages and the stars in each group
    were the same age across temperatures, the velocity dispersion would be
    constant as a function of \teff.
However, the velocity dispersions of the oldest age groups depend on \teff,
    indicating that the \citet{angus2019} gyrochronology model either
    overpredicts the ages of old M dwarfs or underpredicts the ages of old G
    and K dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{age_cut}
\label{fig:age_cut}
\end{figure}
The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample
(excluding visual binaries and subgiants) in grey, with coeval groups shown in
color.
The color of the points corresponds to the age ranges specified in the legend,
which also apply to the lines in the lower panel.
The bottom panel shows the velocity dispersion, \sigmavb\ of each age group,
as a function of effective temperature.
We only included stars within a temperature range of 5000 - 3500 K in our
analysis, as hotter stars are more likely to have stopped magnetic braking
\citep{vansaders2016}, which could bias the results.
M dwarfs were not included in our analysis because such faint stars cannot be
observed at large heights above the plane (because of the low galactic
latitude of the \kepler\ field, stars at high-Z are more distant), which
introduces a mass-dependent velocity bias: cooler populations of stars are
skewed towards lower velocity dispersions.
The coolest temperature bins in figure \ref{fig:age_cut} have low velocity
dispersions, indicating that this effect may already become important at
temperatures lower than $\sim$ 4000 K.

Overall, figure \ref{fig:age_cut} shows that velocity dispersion increases
with gyrochronal age across all temperatures.
However, if the stars in each selected age group had the {\it same} age across
the temperature range, their velocity dispersion would be a constant function
of \teff.
Although the youngest age groups have a relatively constant velocity
dispersions across temperatures, the oldest age groups do not.
This indicates {\it either} that the shape of the period-color relation does
not remain constant over time, \ie\ it flattens out, {\it or} that cool stars
experience more efficient dynamical heating than hot stars.

@@

The consistent velocity dispersion of young stars as a function of temperature
in figure \ref{fig:age_cut} shows that the Praesepe-calibrated gyrochronology
relation accurately predicts the relative ages of {\it young} field stars.
To our knowledge, no gyrochronology relation has never been demonstrated to
correctly predict ages (either relative or absolute) for such cool or such
young field stars.
This is not particularly remarkable however, since these young stars are a
similar age to the Praesepe cluster, which was used to calibrate the
period-color relation.

@@

For now, we assume that the rise in velocity dispersions at cooler effective
temperatures is caused by an inaccurate period-color relation rather than
mass-dependent dynamical heating.
This assumption allows us to estimate what the shape of the period-color
relations {\it should} be.
Figure \ref{fig:age_cut} indicates that the period-color relation flattens
out, so we applied the same analysis to groups of stars with similar rotation
periods, equivalent to a completely flat period-color relation.
The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with stars
in different period ranges plotted in different colors.
The bottom panel shows the velocity dispersion of each group as a function of
effective temperature.
\begin{figure}
  \caption{
This figure is similar to figure \ref{fig:age_cut}, with stars divided into
    period groups rather than age groups.
The velocity dispersion is more constant across effective temperatures for the
    most slowly rotating stars, compared to the stars selected with the
    \citet{angus2019} gyrochronology model, indicating that the gyrochronology
    models flatten out, and possibly even invert, at old ages.
}
  \centering
    \includegraphics[width=1\textwidth]{period_cut}
\label{fig:period_cut}
\end{figure}
Once again, the velocity dispersion increases with rotation period overall.
For the most rapidly rotating groups of stars, velocity dispersion decreases
with \teff\ as expected given the positively sloped period-color relation of
Praesepe and other young clusters: late K dwarfs rotate more slowly than early
K dwarfs of the same age.
Between rotation periods of 15 and 25 days, the temperature dependence of the
velocity dispersion starts to disappear, indicating that the period-color
relation becomes flat: late K dwarfs rotate at the same rate as early K dwarfs
of the same age.
At long rotation periods, the velocity dispersion still increases with
effective temperature, although the increase is much more modest compared to
the oldest stars in figure \ref{fig:age_cut}.
This suggests that rotation period may {\it decrease} with effective
temperature, \ie\ late K dwarfs may rotate {\it more} rapidly than early K
dwarfs.
This would be a paradigm shift for gyrochronology, stellar spin-down rate is
thought to be directly tied to magnetic field strength.
The deeper convection zones of later type stars generate stronger magnetic
fields which {\it should} lead to more efficient angular momentum loss.
Unfortunately, we still cannot rule out the possibility that mass-dependent
heating is acting on these stars which could be responsible for some, if not
all, of the increased velocity dispersion at cooler effective temperatures.

>>>>

\begin{figure}
  \caption{
Top: rotation period vs effective temperature for stars in the \mct\
    catalog.
    The full catalog, with subgiants and visual binaries removed is shown in
    grey, and stars selected to be in different age groups (between 3500 and
    5000 K) are overlayed in color.
These age groups were selected using the \citet{angus2019} gyrochronology
    relation.
The legend in the center of the figure lists the age range, in Gyr, of each
    group.
Bottom: velocity dispersion vs effective temperature for each age
    group.
The color of the line corresponds to the color of the group shown in the top
    panel.
If the gyrochronal model were correct at all ages, and the stars in each group
    were the same age across temperatures, the velocity dispersion would be
    constant as a function of \teff.
However, the velocity dispersions of the oldest age groups increase \teff,
    indicating the \citet{angus2019} gyrochronology model underpredicts the
    the ages of late-K dwarfs relative to the ages of early K dwarfs at old
    ages.
An alternative explanation could be that the gyrochronology relation is
    correct and {\it mass-dependent heating} is responsible for the greater
    velocity dispersions of cooler stars.
}
  \centering
    \includegraphics[width=1\textwidth]{age_cut}
\label{fig:age_cut}
\end{figure}
The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample
(excluding visual binaries and subgiants) in grey, with coeval groups shown in
color.
The color of the points corresponds to the age ranges specified in the legend
(in Gyr), which also apply to the lines in the lower panel.
The bottom panel shows the velocity dispersion, \sigmavb\ of each age group,
as a function of effective temperature.
We only included stars within a temperature range of 5000 - 3500 K in our
analysis, as hotter stars are more likely to have stopped magnetic braking
\citep{vansaders2016}, which could bias the results.
M dwarfs were not included in our analysis because such faint stars cannot be
observed at large heights above the plane (because of the low galactic
latitude of the \kepler\ field, stars at high-Z are more distant), which
introduces a mass-dependent velocity bias: cooler populations of stars are
skewed towards lower velocity dispersions.
The coolest temperature bins in figure \ref{fig:age_cut} have low velocity
dispersions, indicating that this effect may already become important at
temperatures lower than $\sim$ 4000 K.

Overall, figure \ref{fig:age_cut} shows that velocity dispersion increases
with gyrochronal age across all temperatures.
The consistent velocity dispersion of young stars as a function of temperature
shows that the Praesepe-calibrated gyrochronology relation accurately predicts
the relative ages of {\it young} field stars.
% To our knowledge, no gyrochronology relation has never been demonstrated to
% correctly predict ages (either relative or absolute) for such cool or such
% young field stars.
% This is not particularly remarkable however, since these young stars are a
% similar age to the Praesepe cluster, which was used to calibrate the
% period-color relation.
% However, if the stars in each selected age group had the {\it same} age across
% the temperature range, their velocity dispersion would be a constant function
% of \teff.
However, although the youngest age groups have a relatively constant velocity
dispersions across temperatures, the oldest age groups do not.
This indicates {\it either} that the shape of the period-color relation does
not remain constant over time, \ie\ it flattens out, {\it or} that cool stars
experience more efficient dynamical heating than hot stars.

@@

@@

If we assume that the rise in velocity dispersions at cooler effective
temperatures is caused by an inaccurate period-color relation rather than
mass-dependent dynamical heating, we can estimate what the shape of the
period-color relations {\it should} be.
Figure \ref{fig:age_cut} indicates that the period-color relation flattens
out, so we applied the same analysis to groups of stars with similar rotation
periods, equivalent to a completely flat period-color relation.
The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with stars
in different period ranges plotted in different colors.
The bottom panel shows the velocity dispersion of each group as a function of
effective temperature.
\begin{figure}
  \caption{
This figure is similar to figure \ref{fig:age_cut}, with stars divided into
    period groups rather than age groups.
The velocity dispersion is more constant across effective temperatures for the
    most slowly rotating stars, compared to the stars selected with the
    \citet{angus2019} gyrochronology model, indicating that the gyrochronology
    models flatten out, and possibly even invert, at old ages.
}
  \centering
    \includegraphics[width=1\textwidth]{period_cut}
\label{fig:period_cut}
\end{figure}
Once again, the velocity dispersion increases with rotation period overall.
For the most rapidly rotating groups of stars, velocity dispersion decreases
with \teff\ as expected given the positively sloped period-color relation of
Praesepe and other young clusters: late K dwarfs rotate more slowly than early
K dwarfs of the same age.
Between rotation periods of 15 and 25 days, the temperature dependence of the
velocity dispersion starts to disappear, indicating that the period-color
relation becomes flat: late K dwarfs rotate at the same rate as early K dwarfs
of the same age.
At long rotation periods, the velocity dispersion still increases with
effective temperature, although the increase is much more modest compared to
the oldest stars in figure \ref{fig:age_cut}.
This suggests that rotation period may {\it decrease} with effective
temperature, \ie\ late K dwarfs may rotate {\it more} rapidly than early K
dwarfs.
This would be a paradigm shift for gyrochronology, stellar spin-down rate is
thought to be directly tied to magnetic field strength.
The deeper convection zones of later type stars generate stronger magnetic
fields which {\it should} lead to more efficient angular momentum loss.
Unfortunately, we still cannot rule out the possibility that mass-dependent
heating is acting on these stars which could be responsible for some, if not
all, of the increased velocity dispersion at cooler effective temperatures.
